\large How resource abundance and stochasticity affect organisms' space-use requirements

### 2024-03-19 writing/manuscript.Rmd
finished editing discussion

\noindent \textbf{References}: 141 <!-- counted by using nature csl -->

@@

* make sure appendices flow (added stuff from main text)
* check number of references

@@

Examples of ecosystem-level biases are the effects of competition, predation, habitat connectivity, and movement costs, all of which can depend on habitat quality, and, consequently, be correlated nonlinearly to $R$ and $NDVI$ [@jetz_scaling_2004; @prox_framework_2020]. Resource-rich patches can attract larger amounts of competitors [@kacelnik_ideal_1992] and predators, which may increase pressures from competition [@ref] and predation [@brown_ecology_1999]. While such pressures may increase avoidance behaviors, the behaviors may result in both an expansion of the home range [@jetz_scaling_2004; @prox_framework_2020]

***here***

, but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022]. Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [wolves, *Canis lupus*: @rich_anthropogenic_2012; feral cats, *Felis catus*: @bengsen_feral_2016; Capuchin monkeys, *Cebus capucinus*: @torrez-herrera_monkeys_2020], so the effect of $\e(R)$ on space use may vary between individuals, species, and locations. Still, these pressures simply alter the intensity with which organisms respond to $\e(R)$ but not the shape of the relationship, since each of these cases fall along different parts of the continuum in $\e(R)$ shown in fig. \@ref(fig:mu-hyp).

Due to data limitations, disentangling the effects of such confounds is often complex or impossible [e.g., @wang_stochastic_2019].

Satellite-level confounds include coarse spatiotemporal resolution [and the consequent aggregation of signals, see @ref], satellite-level error [@tian_evaluating_2015; @huang_commentary_2021; @fan_global_2016], and other limitations of remote sensing, including the inability to quantify the abundance of specific resources, whether these be preferred forage or the distribution of prey. In the case of preferences for intermediate values of NDVI [e.g., young grass rather than mature grasslands, see @], nonlinear models such as Generalized Additive Models can help account for a nonmonotonic relationship between $H$ and $E(NDVI)$, as long as data are sufficiently abundant to avoid any artifacts due to data paucity. In the case study we present here, sampling limitations resulted in a strong nonlinear correlation between $\e(R)$ and $\var(R)$, which prevented us from estimating any nonlinear preferences in mean NDVI (i.e., the proxy for $\e(R)$). A tracking time series of two or more years or additional individuals would have provided better estimates, but that was beyond the scope of this paper.

Modeling the tapir's space use as a function of resource abundance alone would result in an excessively strong estimated effect and a quantifiably worse model fit ($\Delta\text{AIC} \approx 87$, see Appendix C). By including the effect of $\sigma^2(t)$ in the model along with that of $\mu(t)$, we show that the tapir responds to both resource abundance and stochasticity. While the tracking period may be too short to see the effects of $\e(R)$ clearly, it should also be noted that the Cerrado is relatively rich in vegetation throughout the entire year (fig. \@ref(fig:tapir-mw)A), and thus the tapir does not experience resource scarcity during the tracking period. Consequently, this example demonstrates two points central to this paper. Firstly, even animals in resource-rich regions such as the Brazilian Cerrado adapt their behavior in response to reosurce stochasticity. Secondly, trends in mean resources alone are not sufficient to explain the tapir's space use, and the stochasticity in $R$ is an important driver of the tapir's movement.

## Adaptations to changes in $\e(R)$ and $\var(R)$

@@

<!-- spores are also an adaptation to periods of low $\e(R)$ or high $\var(R)$ -->

\noindent The spatiotemporal scale over which an event occurs is a main determinant of whether an organism will be able to predict the event and how it will respond to it (or fail to do so). Events that occur at the scale of the organism [i.e. approximately one or more times per generation and on the spatial scale of the organism's spatial range, see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [@foley_severe_2008; @berger_climate_2018; @geremia_migrating_2019]. Consequently, organisms with long generations (e.g., elephants, *Loxodonta sp.*) or large spatial ranges (e.g., migratory birds) are more likely to experience stochastic events and thus are also more likely to respond to and prepare for them. Animals may adapt by adjusting the timing of reproduction [@southwood_habitat_1977; @boersma_wildfire_2021] and torpor or hibernation [@inouye_climate_2000; @goldberg_hibernation_2021], or storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [@armitage_energetics_2003; @nespolo_why_2022] or resource caching [@post_caching_2006]. Alternatively, animals may alter their movement and travel long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017], whether this be for migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. Over the years, animals may also develop more complex cognitive abilities [@brown_toward_2004], including the development of memory [@fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; @geremia_migrating_2019; @abrahms_memory_2019; @cavedon_genomic_2022; @foley_severe_2008; @polansky_elucidating_2015; @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; @ranc_memory_2022], which is required for migration [@mueller_social_2013; @geremia_migrating_2019; @merkle_spatial_2019; @abrahms_memory_2019; @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].
<!-- 
with species names:
Organisms may adapt by adjusting the timing of reproduction [framework: @southwood_habitat_1977; red-backed fairywren, *Malurus melanocephalus*: @boersma_wildfire_2021] and torpor or hibernation [marmots, *Marmota flaviventris*: @inouye_climate_2000; ground squirrels, *Urocitellus brunneus*: @goldberg_hibernation_2021], storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [marmots: @armitage_energetics_2003; bears, *Ursus sp.*: @nespolo_why_2022] or resource caching [@post_caching_2006]. Alternatively, animals may alter their movement and travel long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017], whether this be for migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. Over the years, animals may also develop more complex cognitive abilities [@brown_toward_2004], including the development of memory of events [framework: @fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; bison and mule deer: @geremia_migrating_2019; blue whales, *Balaenoptera musculus*: @abrahms_memory_2019; caribou, *Rangifer tarandus*: @cavedon_genomic_2022; elephants: @foley_severe_2008; @polansky_elucidating_2015; elk, *Cervus canadensis*: @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; roe deer, *Capreolus capreolus*: @ranc_memory_2022], which is required for migration [whooping cranes, *Grus americana*: @mueller_social_2013; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].-->
However, adapting to changes in environmental cues rather than relying on memory alone [@severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].

@@

\noindent The work presented here provides a unifying framework for viewing movement as a function of resource abundance and stochasticity. We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on organisms' space-use requirements and movement behavior. We demonstrate that organisms' space-use requirements respond nonlinearly to both resource abundance and stochasticity, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$. We also provide a moving window approach that allows one to model movement continuously rather than imposing arbitrary thresholds between different periods or types of behaviors, including temporal changes in space-use requirements and shifts between range-residency and nomadism. The methods presented here can be applied to assess the current and future quality of protected areas while also estimating the amount of space organisms currently need and will need in the future under different climate change scenarios.

>>>>

\noindent \textbf{References}: 95 <!-- counted by using nature csl -->

@@

* make sure appendices flow (added stuff from main text)
* check number of references (updated on 2024-03-18)

@@

Examples of ecosystem-level biases are the effects of competition, predation, habitat connectivity, and movement costs, all of which can depend on habitat quality, and, consequently, be correlated nonlinearly to $R$ and $NDVI$ [@jetz_scaling_2004; @prox_framework_2020]. Resource-rich patches can attract larger amounts of competitors [@kacelnik_ideal_1992] and predators, which may increase pressures from competition [@ref] and predation [@brown_ecology_1999]. While such pressures may increase avoidance behaviors, avoidance may have a cmplex effect on $H$, since it can result in both an expansion of the home range [@jetz_scaling_2004; @prox_framework_2020] or a contaction of an animal's movement, since larger home ranges can be harder to defend and also result in higher movement costs [@grant_whether_1993; @jetz_scaling_2004]. Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [theoretical framework: @despotic_ref; examples: @rich_anthropogenic_2012; @bengsen_feral_2016; @torrez-herrera_monkeys_2020].

Satellite-level confounds include information loss due to coarse spatiotemporal resolution [@ref], satellite-level error [@tian_evaluating_2015; @huang_commentary_2021; @fan_global_2016], and other limitations of remote sensing (e.g., inability to quantify specific resources, competitive pressure, and predator avoidance). However, nonlinear models such as Generalized Additive Models [@wood_generalized_2017] can help account for preferences for intermediate values of NDVI [e.g., young grass rather than mature grasslands, see @ref]. In the case study we present here, sampling limitations resulted in a strong nonlinear correlation between $\e(R)$ and $\var(R)$ that prevented us from estimating nonlinear preferences in mean NDVI (i.e., the proxy for $\e(R)$), since estimating the divers of movement and disentangling them from potential confounds in a stochastic environemtn is often complex or impossible [e.g., @wang_stochastic_2019]. A tracking time series of two or more years or additional individuals would have provided better estimates, but that was beyond the scope of this empirical example.

**here:** need a few lines to tie this back to the general

<!-- Feel free to delete this: -->
<!-- While the tracking period may be too short to see the effects of $\e(R)$ clearly, it should also be noted that the Cerrado is relatively rich in vegetation throughout the entire year (fig. \@ref(fig:tapir-mw)A), and thus the tapir does not experience resource scarcity during the tracking period. Consequently, this example demonstrates two points central to this paper. Firstly, even animals in resource-rich regions such as the Brazilian Cerrado adapt their behavior in response to reosurce stochasticity. Secondly, trends in mean resources alone are not sufficient to explain the tapir's space use, and the stochasticity in $R$ is an important driver of the tapir's movement. -->

<!-- 
## Adaptations to changes in $\e(R)$ and $\var(R)$

@@

The spatiotemporal scale over which an event occurs is a main determinant of whether an organism will be able to predict the event and how it will respond to it (or fail to do so). Events that occur at the scale of the organism [i.e. approximately one or more times per generation and on the spatial scale of the organism's spatial range, see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [@foley_severe_2008; @berger_climate_2018; @geremia_migrating_2019]. Consequently, organisms with long generations (e.g., elephants, *Loxodonta sp.*) or large spatial ranges (e.g., migratory birds) are more likely to experience stochastic events and thus are also more likely to respond to and prepare for them. Animals may adapt by adjusting the timing of reproduction [@southwood_habitat_1977; @boersma_wildfire_2021] and torpor or hibernation [@inouye_climate_2000; @goldberg_hibernation_2021], or storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [@armitage_energetics_2003; @nespolo_why_2022] or resource caching [@post_caching_2006]. Alternatively, animals may alter their movement and travel long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017], whether this be for migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. Over the years, animals may also develop more complex cognitive abilities [@brown_toward_2004], including the development of memory [@fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; @geremia_migrating_2019; @abrahms_memory_2019; @cavedon_genomic_2022; @foley_severe_2008; @polansky_elucidating_2015; @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; @ranc_memory_2022], which is required for migration [@mueller_social_2013; @geremia_migrating_2019; @merkle_spatial_2019; @abrahms_memory_2019; @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].
-->

<!-- 
with species names:
Organisms may adapt by adjusting the timing of reproduction [framework: @southwood_habitat_1977; red-backed fairywren, *Malurus melanocephalus*: @boersma_wildfire_2021] and torpor or hibernation [marmots, *Marmota flaviventris*: @inouye_climate_2000; ground squirrels, *Urocitellus brunneus*: @goldberg_hibernation_2021], storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [marmots: @armitage_energetics_2003; bears, *Ursus sp.*: @nespolo_why_2022] or resource caching [@post_caching_2006]. Alternatively, animals may alter their movement and travel long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017], whether this be for migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. Over the years, animals may also develop more complex cognitive abilities [@brown_toward_2004], including the development of memory of events [framework: @fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; bison and mule deer: @geremia_migrating_2019; blue whales, *Balaenoptera musculus*: @abrahms_memory_2019; caribou, *Rangifer tarandus*: @cavedon_genomic_2022; elephants: @foley_severe_2008; @polansky_elucidating_2015; elk, *Cervus canadensis*: @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; roe deer, *Capreolus capreolus*: @ranc_memory_2022], which is required for migration [whooping cranes, *Grus americana*: @mueller_social_2013; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].-->
<!-- However, adapting to changes in environmental cues rather than relying on memory alone [@severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].

@@

**needs a few more lines at the end**

\noindent The work presented here provides a unifying framework for viewing movement as a response to resource abundance and stochasticity. We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on organisms' space-use requirements and movement behavior. We demonstrate that organisms' space-use requirements respond nonlinearly to both resource abundance and stochasticity, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$. We also provide a moving-window approach that allows one to model movement continuously rather than imposing arbitrary thresholds between different periods or types of behaviors, including temporal changes in space-use requirements and shifts between range-residency and nomadism. The methods presented here can be applied to assess both current and future quality of protected areas while also estimating the amount of space organisms currently need and will need in the future under different climate change scenarios.
